Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices

### 2022-08-07 supp/index.Rmd
<edit index>

--- 
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices and raw material differences"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade and/or exchange with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent due to idiosyncratic responses related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences and reduction practices. The analysis of biface morphlogy by size class demonstrates that size does not differ between raw materials in the large class, but does in the small class. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife. In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache. Distinct reduction practices and raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric). The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

@@

Landmark data were subsequently aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v. 4.0.4 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

>>>>

--- 
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood object from the ancestral Caddo area. Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and thought to have been acquired by the Caddo through trade and/or exchange with central Texas groups. Bifaces from the Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical assessment of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces from each site was divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to resharpening and/or retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential. The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class. This finding supports the argument that elliptical biface morphology diverges through local resharpening and/or retouch practices, where bifaces manufactured using both raw materials enter their uselife at similar shapes and sizes prior to diverging in size. As expected, greater shape diversity occurs in the small class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from the Millsap Cache. Distinct reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric). The number of flake scars was then divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

@@

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v. 4.0.4 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space following alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
